---
title: "CTaylorMSDS6306402Unit6CaseStudy GDP Vs EDSTAT"
author: "Celia Taylor"
date: "June 21, 2016"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Case Study of Country GDP and Country Federal Statistics
##### This study is to answer 5 main questions.

##### Question 1	Match the data based on the country shortcode. How many of the IDs match between the data from GDP data file and EDSTAT data file?

#####Question 2	Sort the data frame in ascending order by GDP rank (so United States is last). What is the 13th country in the resulting data frame?

#####Question 3	What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?

#####Question 4	Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.

#####Question 5	Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?

#####To answer these questions and learn something about the economics of the countries of the world, the specific data from two different files will be brought together, analyzed, and appropriately handled to show the data.
#####All steps are shown for the integrity of the case study.

#####Download data files, specifically Federal Gross Domestic Product information (GDP) and Federal Statistics by Country (EDSTAT) information.


```{r dwnldrCSV1,  include=TRUE, echo=TRUE}
install.packages("downloader", repos="http://cran.rstudio.com/")
library(downloader)
download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv", destfile="GDP.csv")
download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv", destfile="edstat.csv")
#list.files()
```


```{r Install_packages_1, include=TRUE, echo=TRUE}
##### Load packages and functions

#install.packages(c( "plyr",  "brew",  "countrycode",  "devtools", "dplyr", "ggplot2", "googleVis", "knitr", "MCMCpack", "repmis", "RCurl", "rmarkdown", "markdown", "texreg", "tidyr", "WDI", "xtable", "Zelig", "car", "Rtools"), repos="http://cran.rstudio.com/")

#library(plyr)
#library(brew)
#library(countrycode)
#library(devtools)
#library(ggplot2)
#library(googleVis)
#library(knitr) 
#library(MCMCpack) 
#library(repmis) 
#library(RCurl) 
#library(rmarkdown) 
#library(texreg)
#library(tidyr) 
#library(WDI) 
#library(xtable) 
#library(Zelig) 
#library(markdown)
#library(car)
#library(dplyr)
#sessionInfo()
#End of install packages
library(ggplot2)
library(dplyr)
```
##### Read in edstat.csv file with additional options.
```{r readEDcsv1, include=TRUE, echo=TRUE}
#####
##### Read in edstat.csv file with additional options.  Get rid of any blank lines.
ed <- read.csv("edstat.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE, nrows = 236, blank.lines.skip = TRUE)
str(ed)
#Make a backup copy to manipulate
edraw <- ed
#Make sure structure is okay
str(edraw)
head(edraw,2)
tail(edraw,2)

```
##### Read in GDP.csv file with additional options.
```{r readGDPcsv1, include=TRUE, echo=TRUE}
##### Read in GDP.csv file with additional options.
gdp <- read.csv("GDP.csv", header = FALSE, sep = ",", stringsAsFactors=FALSE, nrows = 236, blank.lines.skip = TRUE)

#Make a backup copy to manipulate
gdpraw <- gdp
#Make sure structure is okay
str(gdpraw)
head(gdpraw,2)
tail(gdpraw,2)

```
##### Keep only the columns in edraw that are needed for analysis
```{r cleanEdCol1, include=TRUE, echo=TRUE}
##### Keep only the columns in edraw that are needed for analysis
edraw1 <- edraw[c(1,3,31)]
str(edraw1)
```
##### Keep only the columns in gdpraw that are needed for analysis
```{r cleanGDPCol1, include=TRUE, echo=TRUE}
##### Keep only the columns in gdpraw that are needed for analysis
gdpraw1 <- gdpraw[c(1,2,4,5)]
str(gdpraw1)
```
##### Rename headers in edraw to good and descriptive names without spaces
```{r cleanEdHead1, include=TRUE, echo=TRUE}
###### Rename headers in edraw to good and descriptive names without spaces
edraw2 <- edraw1
names(edraw2) <- c("EdCountryCode", "IncomeGroup", "ShortName")
str(edraw2)
```
##### Rename headers in gdpraw to good and descriptive names without spaces
```{r cleanGDPHead1, include=TRUE, echo=TRUE}
##### Rename headers in gdpraw to good and descriptive names without spaces
str(gdpraw1)
head(gdpraw1, 2)
#Get rid of top rows that do not have good information
gdpraw2 <- gdpraw1[6:236,]
names(gdpraw2) <- c("GdpCountryCode", "Ranking", "Economy", "MillionsOfDollars")
str(gdpraw2)
head(gdpraw2, 2)
tail(gdpraw2, 2)
```

##### Keep only the rows in gdpraw1 that are needed for analysis
```{r cleanGDPRow1, include=TRUE, echo=TRUE}
#Lines 6 through 195 have good information on Ranking and MillionsOfDollars, 
#so make a data frame that concentrates on that data with the added benefit of
#eliminating extraneous and possible troublesome (NAs) data.
gdpAnal1 <- gdpraw1[6:195,]
names(gdpAnal1) <- c("GdpCountryCode", "Ranking", "Economy", "MillionsOfDollars")
str(gdpAnal1)
#Force "Ranking" to be an integer for easier sorting
gdpAnal1$Ranking <- as.integer(gdpAnal1$Ranking)
#Check structure and make sure integer conversion worked.
str(gdpAnal1)
head(gdpAnal1, 2)
tail(gdpAnal1, 2)

```

#####Question 1	Match the data based on the country shortcode. How many of the IDs match? 
##### Merge data frames by CountryCode
```{r MergeCountryCode1, include=TRUE, echo=TRUE}
#This is the data frame that has the larger set to find all matching country shortcode IDs from the #from column 1 of GDP data and column 1 of the edstats data.
MergeEdGdp4 <- merge(edraw2, gdpraw2, by.x = "EdCountryCode", by.y = "GdpCountryCode", sort = TRUE, all = FALSE, incomparables = NULL)
str(MergeEdGdp4)
head(MergeEdGdp4, 2)
tail(MergeEdGdp4, 2)
```
##### There are 224 matching IDs or country shortcodes.  This includes regional or economic country shortcode IDs.

#####Question 2	Sort the data frame in ascending order by GDP rank (so United States is last). What is the 13th country in the resulting data frame?
```{r SortAscGDPRank1, include=TRUE, echo=TRUE}
#Utilize that gdpAnal1 data frame that has the 190 countries with good Ranking 
#and MillionsOfDollars
#SSD - South Sudan doesn't match because not in edstat data.
SortGdpRank189 <- merge(edraw2, gdpAnal1, by.x = "EdCountryCode", by.y = "GdpCountryCode", sort = TRUE, all = FALSE, incomparables = NULL)
#Keep SSD - South Sudan in data frame to utilize Ranking and MillionsOfDollars data.
SortGdpRank <- merge(edraw2, gdpAnal1, by.x = "EdCountryCode", by.y = "GdpCountryCode", sort = TRUE, all.y = TRUE, incomparables = NULL)
#Make sure structure is OK and all 190 countries are there.
str(SortGdpRank189)
str(SortGdpRank)
head(SortGdpRank, 2)
tail(SortGdpRank, 2)
#Show SSD South Sudan as list number 155 with NA in IncomGroup.
SortGdpRank
#Manipulate a backup copy
SortGdpRank1 <- SortGdpRank
#
SortGdpRank2 <- SortGdpRank1[order(SortGdpRank1$Ranking, decreasing = TRUE), ]
#Make sure structure is ok and show the first 13 on list and that USA is last. 
str(SortGdpRank2)
#Show the first 13 countries in newly ordered data frame.
head(SortGdpRank2, 13)
#Show that USA is last with rank #1.
tail(SortGdpRank2, 1)

```
##### There are 190 countries from the GDP table that have a valid ranking.  When sorted from lowest ranking to the highest ranking (USA), the 13th country in the sorted data frame is St. Kitts and Nevis.  St. Kitts and Nevis has a tied Ranking to Grenada.
########  Please Note there is no information for SSD South Sudan (155 on list) in the edstat data but there is data in the GDP data. South Sudan was formed July 9, 2011. The data on this country that is not available is the Income Group Data.  According to http://www.worldbank.org/en/country/southsudan/overview#1, "The country’s growth domestic product (GDP) per capita in 2014 was $1,111. Outside the oil sector, livelihoods are concentrated in low productive, unpaid agriculture and pastoralists work, accounting for around 15% of GDP. In fact, 85% of the working population is engaged in non-wage work, chiefly in agriculture (78%)."

 
#####Question 3	What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups? 
```{r AverageHiInc1, include=TRUE, echo=TRUE}
#Work with backup copy and use data frame without South Sudan
SortGdpRank3 <- SortGdpRank189
#Collect the rows into a new data frame that have "High income: OECD" in the IncomeGroup column
HighIncOECD <- SortGdpRank3[SortGdpRank3$IncomeGroup == "High income: OECD", c(1,2,3,4,5,6) ]
#Collect the rows into a new data frame that have "High income: nonOECD" in the IncomeGroup column
HighIncNonOECD <- SortGdpRank3[SortGdpRank3$IncomeGroup == "High income: nonOECD", c(1,2,3,4,5,6)]
#Make sure structure is ok and find out how many are in each group.
str(HighIncOECD)
HighIncOECD
str(HighIncNonOECD)
HighIncNonOECD

```

##### The average GDP rankings for the "High income: OECD".  

######## Data processed without South Sudan information.

```{r AverageHiInc2, include=TRUE, echo=TRUE}
AvGDPRankOECD <- mean(HighIncOECD$Ranking)
AvGDPRankOECD
```
##### The average GDP rankings for the "High income: nonOECD"
```{r AverageHiInc3, include=TRUE, echo=TRUE}
AvGDPRankNonOECD <- mean(HighIncNonOECD$Ranking)
AvGDPRankNonOECD
```
######## Data processed without South Sudan information.

#####Question 4	Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.

#####   GDP FOR ALL COUNTRIES
```{r PlotIncGroup1, include=TRUE, echo=TRUE}
#### grouped box plots
#####Take out the commas in the MillionsOfDollars
SortGdpRank3 <- SortGdpRank189
SortGdpRank3$MillionsOfDollars <- gsub(",", "", SortGdpRank3$MillionsOfDollars)
SortGdpRank3$MillionsOfDollars <- gsub(",", "", SortGdpRank3$MillionsOfDollars)

#Force "MillionsOfDollars" to be an integer
SortGdpRank3$MillionsOfDollars <- as.integer(SortGdpRank3$MillionsOfDollars)
str(SortGdpRank3)
tail(SortGdpRank3, 20)
SortGdpRank4 <- SortGdpRank3
SortGdpRank4$MillionsOfDollars <- (SortGdpRank4$MillionsOfDollars/1000000)
str(SortGdpRank4)
tail(SortGdpRank4, 20)
```

#####   GDP FOR ALL COUNTRIES - Plotted in Color with Y scale in millions
```{r PlotIncGroup2, include=TRUE, echo=TRUE}
ggplot(SortGdpRank3,  aes(SortGdpRank3$IncomeGroup, SortGdpRank3$MillionsOfDollars, fill=SortGdpRank3$IncomeGroup )) +  geom_point() + geom_boxplot()
```

#####   GDP FOR ALL COUNTRIES  - Plotted in Grey with Y scale divided by millions
```{r PlotIncGroup3, include=TRUE, echo=TRUE}
ggplot(SortGdpRank4,  aes(SortGdpRank4$IncomeGroup, SortGdpRank4$MillionsOfDollars )) +  geom_point() + geom_boxplot()


```














#####The USA is such an outlier at 16,244,600 million that the scale is hard to adjust to show the other boxplots' color or grey.
######## Data processed without South Sudan information.

#####Question 5	Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?

```{r QuantTableIncGroup1, include=TRUE, echo=TRUE}

########5	Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?


QuantGdpRank3 <- SortGdpRank2
QuantGdpRank3$quantile <- ntile(QuantGdpRank3$Ranking, 5)
str(QuantGdpRank3)
QuantGdpRank3
table(QuantGdpRank3$quantile,QuantGdpRank3$IncomeGroup)

```
##### The answer is 5 countries are Lower Middle Income but among the 38 nations with the highest GDP.  The countries are China, India, Indonesia, Thailand, and Egypt.  This data is processed without South Sudan information.

#################################################################################################

##### There were 224 matching country codes.  After processing the matching codes, I then merged only the 190 rows from the GDP data set that had rankings and Millions of Dollars in GDP, so I utilized only the needed information and did not keep working with potentially disruptive data.  So potentially, there were 34 rows that had NA or disruptive data.  Additionally, South Sudan was not in the edstat data, because it is such a new country.  Because there was a lack of data about South Sudan's IncomeGroup, the calculations or graphics that included IncomeGroup were done with 189 countries instead of 190.

#################################################################################################
#################################################################################################
#################################################################################################
#################################################################################################
#################################################################################################
########Document the installed packages and libraries

```{r InstallPackLib1, include=TRUE, echo=TRUE}
sessionInfo()

```





